A hierarchical approach for modelling X-ray beamlines. Application to a coherent beamline.

### 2019-06-14 paper-hierarchical.tex
Rafael changes

\begin{synopsis}
A hierarchical scheme of the calculations of an X-ray beamline for coherent applications is presented. Starting from simple calculations to partial coherence calculations, we show how ray tracing and wave optics software are applied simulating the beamline performances.   
\end{synopsis}

\begin{abstract}
We consider different approaches to simulate a modern X-ray beamline. Several methodologies with increasing complexity are applied to discuss the relevant parameters that quantify the beamline performance. Parameters such as flux, dimensions and intensity distribution of the focused beam and coherence properties are obtained from simple analytical calculations to sophisticated computer simulations using ray tracing and wave optics techniques. A latest generation x-ray nanofocusing beamline for coherent  applications (ID16A at the ESRF) has been chosen  to study in detail the issues related to highly demagnifying  synchrotron sources and exploiting the beam coherence. The performance of the beamline is studied for two storage rings: the old ESRF-1 (emittance 4000 pm) and the new ESRF-EBS (emittance 150 pm). In addition to traditional results in term of flux and beam sizes, an innovative study on the partial coherence properties based on propagation of coherent modes is presented. The different algorithms and methodologies are implemented in the software suite OASYS \cite{codeOASYS}. They are discussed with emphasis beamline placed upon the benefits and limitations of each. 
\end{abstract}

@@

Many storage ring based x-ray synchrotron facilities are building or planning upgrades to increase brilliance and coherent flux by one to three orders of magnitude.  The first upgrade of a large facility will be the EBS (Extremely Brilliant Source) \cite{orangebook} at the European Synchrotron Radiation Facility (ESRF). The new storage ring of 150 pm emittance that will be commissioned in 2020 will significantly boost the coherence of the x-ray beams. Applications exploiting beam coherence such as x-ray photon correlation spectroscopy, coherent diffraction imaging, propagation-based phase contrast imaging and ptychography will strongly benefit from this update.

Accurate calculation and quantitative evaluation of the parameters related to x-ray optics is of great importance for designing, building and exploiting new beamlines. Every modern beamline follows a procedure of conceptual design plus technical design where detailed simulations are essential. These calculations  allowing testing of the design parameters in a virtual computer environment where advantages and limitations can be studied quantitatively. Selection of the optimal optical layout requires the detailed simulation of the optics imperfections, that in many cases are the limiting factor of the beamline optical configuration. Deformations of the optical elements due to heat load need to be controlled, therefore optics simulations have to include results from engineering modeling of the thermal deformations usually done using finite element analysis. Traditionally the optics calculations for synchrotron beamlines are done using ray tracing techniques. They model the x-ray beam as a collection of incoherent rays. This technique is still essential for obtaining information on aberrations, flux propagation, monochromator bandwidth, etc. However, the high coherence of the new sources can introduce the necessity to complement the ray tracing methods with models including effects of diffraction and scattering from a coherent beam. A fully coherent hard x-ray beam is still a dream for the new generation storage rings. In spite of a considerable increase in the coherent fraction of the new beams, the coherent fraction at typically hard x-ray photon energies is still a few percent for the new sources. Consequently ray tracing analysis does not become obsolete, but has to be complemented with an analysis based on wave optics. Because of the partial coherence the wave optics analysis becomes non trivial and can be done with different levels of approximation, as discussed in this work. In addition, wave optics simulations are usually computationally expensive.

@@

X-ray beamlines are very particular optical systems. Some implications of exploiting the short wavelengths such as the use of grazing incidence reflectors and the use of perfect crystals as typical hard x-ray monochromators render common commercial software packages ill-adapted for the study of synchrotron beamlines. For this reason many optics tools have been developed by the synchrotron community. Some of them have a long history like SHADOW (more than 40 years old) or SRW (almost 30), to mention only codes used in this work. These historical tools together with other new applications have been collected together in a user friendly software package: OASYS. Many of the calculations presented here have been performed directly in OASYS or with scripts created with the help of OASYS.

The paper is structured as follows: we first introduce in Section~\ref{Beamline description} the main parameters of the beamline and explain the particular choice of the parameters used in the calculations. Section~\ref{Simulations} is the central part of the paper, and describes the different methods used to analyse the beamline with increasing complexity, setting a hierarchy of methodologies that can be exported to the simulations of any beamline. In Section \ref{level0} we calculate with simple methods approximate values of focal size, flux and coherent fraction. Accuracy can improved using geometrical methods in Section \ref{level1} that allow a more complete simulation of the beamline optics within certain limits which are related, in this example, to high demagnification and high coherence of the studied beamline. The wave optics section~\ref{level2} includes different models of increasing complexity as the required accuracy of the results becomes more demanding. In the Discussion (Section \ref{discussion}) the results of the different simulation methods are commented. The hierarchical approach allows more accurate and better results but the price to pay is to deal with more complex and computationally expensive tools. Finally Section \ref{summary} summarizes the results.

@@

This section presents the reader with the main operational parameters for the ID16A beamline at ESRF in two scenarios. This beamline was built during the Upgrade Phase I at ESRF (2009-2015). It uses one or several undulators as source. During the period 2011-2018 it operated exploiting the ESRF-1 storage ring’s double-bend achromat (DBA) lattice with horizontal emittance about 4000 pm. In the second phase of the ESRF Upgrade the original storage ring is being replaced by a new hybrid multi-bend achromat (HMBA) lattice. This new lattice will drastically improve the performances of the ESRF source. The resulting Extreme Brilliant Source (EBS) \cite{orangebook} will increase the brightness and coherent fraction by a factor of more than 100, by strongly reducing the horizontal emittance. This is possible using almost the same ring circumference and electron energy $E_e$. As a result, the x-ray sources will be much coherent, specially in the horizontal direction. In this paper, some quantitative results on how the storage ring upgrade will affect a particular beamline are presented. 
%RC190404: @PhaseI -> from the orange book: On the basis of the Purple Book, the ESRF Council decided in June 2008 to launch the ESRF UP by establishing a two-phase process. The Upgrade Programme Phase I (UP PI) for the period 2009-2015, and the Upgrade Programme Phase II (UP PII) to be prepared in the period 2012-2014, and to be presented to the Council for consideration in 2014 and launch in 2015.
% The  Upgrade  Programme  Phase  I  has  been  on  the  ESFRI  Roadmap  since  its  fi rst publication  in  2006,  and  is  today  heading  towards  its  successful  completion  by  the end of 2015.

ID16A \cite{ID16A} is a beamline that provides a nanometric, highly coherent beam with broad energy-bandpass. It is dedicated to nanofocusing applications. The beamline can operate at two fixed photon energies: 17050 eV and 33600 eV. The straight section in the magnet lattice houses four undulators, two 18 mm and another two of 22.4 mm period. The undulator parameters and $K$-values at the resonances for the energies in use are shown in Table \ref{tab:Undulators}.

%RC190404: From the ID16D page -> The beamline provides a high-brilliance beam focused down to nanometer size, allowing quantitative 3D characterization of the morphology and the elemental composition of specimens in their native state by combining coherent imaging techniques and X-ray fluorescence microscopy.
% The 185 m long beamline provides nano-focused beams for analytical imaging. The nano-imaging beamline addresses problems in biology, biomedicine and nano-technology using X-ray fluorescence microscopy and nano-tomography. It is optimised for ultimate hard X-ray focusing of a beam with a large energy bandwidth at specific energies (17keV or 33.6 keV). Cryogenic sample preservation is available for 2D and 3D imaging.
% Techniques: coherent imaging; phase-contrast imaging; near-field and far-field ptychography; X-ray fluorescence microscopy;  scanning transmission X-ray microscopy; nanotomography
% http://www.esrf.eu/home/UsersAndScience/Experiments/XNP/ID16A/over.html

\begin{table}\label{tab:Undulators}
\centering
\caption{Parameters of the undulators available at the ID16A beamline: period ($\lambda_u$), length ($L$), number of identical undulators ($N$) and deflection parameter ($K$) values for the three main photon energies in use. The valuse are shown for the two ESRF magnetic lattices: the old ESRF-1 (labelled ESRF, with $E_e = 6.04$ GeV) and the new ESRF-EBS (labelled EBS, with $E_e = 6.00$ GeV). $n$ is the emission harmonic in use for each particular photon energy.}

\resizebox{\textwidth}{!}{%
\begin{tabular}{lcccccc}
name  & $\lambda_u$ [mm] & $L$ [m] & N & $K$ (11200 eV) & $K$ (17050 eV) & $K$ (33600 eV) \\ 
\input{tableUndulator.txt}
\end{tabular}%
}
\end{table}

ID16A is a long beamline (the distance from the center of the straight section to the sample position is 185 m). Strong focusing is performed using a Kirkpatrick-Baez (KB) mirror system placed a few cm upstream of the sample position (focal plane), which ensures a high demagnification. In order to match the acceptance of the KB in the vertical and horizontal planes a focusing (cylindrically bent) multilayer monochromator (ML) has
been chosen. It i) monochromatises the beam with a large energy bandwidth (typically $\Delta E/E \approx 10^{-2}$) to allow high flux, 
and ii) it focus tangentially the beam  in the horizontal plane to create a secondary secondary source at 40 m downstream of the undulator, where a horizontal slit (VSS) is placed. A beamline schematic is shown in Figure~\ref{fig:ID16A}. The positions of the different elements are shown in Table \ref{tab:Positions}. This table also shows the extremely high geometrical demagnification ($\sim 10^3$) needed for focusing the beam to the nanometric range.

@@

Simulations were done using the two different storage ring lattices mentioned before: the (already dismantled) ESRF-1 where the undulators are in a high-$\beta$ straight section (larger horizontal emittance) and the new EBS (lower horizontal emittance) straight sections. The parameters for the current and the new lattice are shown in Tables \ref{tab:TableSources} and \ref{tab:eBeam}.

\begin{table}\label{tab:TableSources}
    \caption{ESRF and ESRF-EBS main storage ring parameters. Main parameters of the present DBA lattice (ESRF-1) and future HMBA lattice (EBS) storage rings \cite{orangebook}.}
        \begin{tabular}{rcc}
                                              & \textbf{ESRF-1} & \textbf{EBS}      \\ 
        Lattice type:                         & DBA           & HMBA              \\
        Circunference {[}m{]}:                & 844.930       & 843.979           \\
        Beam energy {[}GeV{]}:                & 6.04          & 6.00              \\
        Beam current {[}mA{]}:                & 200           & 200               \\
        Natural emittance {[}pm$\cdot$rad{]}: & 4000          & 147               \\
        Energy spread {[}$\%${]}:             & 0.0011        & 0.00093           \\       
    \end{tabular}
\end{table}

\begin{table}\label{tab:eBeam}
    \centering
    \caption{ESRF-1 and EBS main electron beam parameters. Electron beam parameters for the high-$\beta$ (labelled ESRF) and the EBS straight sections. Values taken at the symmetry point of the straight section, where the insertion devices are placed. Values for ESRF are taken from \cite{ESRF2014}. For EBS, two (very close) design values are used, from Ref.~\cite{ESRF2014} labelled EBS-S28A, and those coming from further lattice refinement, labelled EBS-S28D. \todo{Remove ESRF-AT and EBS-S28A} }
        \resizebox{\textwidth}{!}{%
        \begin{tabular}{ccccc}
         Lattice & $\sigma_e$ horizontal {[}$\mu$m{]} &  $\sigma_e'$ horizontal {[}$\mu$rad{]} & $\sigma_e$ vertical {[}$\mu$m{]} & $\sigma_e'$ vertical {[}$\mu$rad{]} \\
         \input{tableElectron.txt}

@@

For all simulations presented hereafter, we selected the configuration using a single undulator U18.3 placed in the center of the straight section and tuned to its first harmonic at a photon energy of $E=17225$ eV (a value close to the usual operation value shown in Table \ref{tab:Undulators}). This section is divided in three subsections, that present three main levels of simulations: simple calculations that can be made by hand (section \ref{level0}), ray-tracing simulations (section \ref{level1}) and wave optics simulations (section \ref{level2}).

@@

We first estimate the beam size at different positions in the beamline, in particular at the source plane and at the image plane. A first na{\"{i}}ve estimation of the focal size can be made by considering the source size times the magnification factor $M$. In the paraxial approximation, the magnification factor can be expressed as $M=q/p$, where $p$ is the distance between object and imaging element and $q$ is the distance between imaging element and image. The demagnification is the inverse of the magnification ratio: $M^{-1}=p/q$. Systems where $M>1$ are said to be magnifying, whereas systems with $M<1$ (or $M^{-1}>1$) are said to be demagnifying. The latter will be our case, thus is more convenient to work with demagnification ($M^{-1}>1$) rather than with magnification.

The source size (or divergence) is usually approximated supposing a Gaussian distribution of the electron positions (or divergence) (Table \ref{tab:eBeam}) and single-electron photon source size or divergence. In some cases, such as for undulator emission, this Gaussian approximation may not be accurate and have some implications that will be discussed later. Let $\sigma_{e(h,v)}$ and $\sigma_{e(h,v)}'$ be the standard deviations of electron beam size and divergence  at the centre of  symmetry of the straight section in a plane perpendicular to the main propagation direction. The subscripts $h$ and $v$ stand for horizontal and vertical, respectively.

The undulator natural photon source size and divergence are given by \cite{elleaume}:
\begin{align}
    \label{eq:photon small sigmas}
    \sigma_u=\frac{2.74}{4\pi}\sqrt{\lambda L}\approx \frac{1}{2  \pi}\sqrt{2 \lambda L}  && \sigma_u' = 0.69\sqrt{\frac{\lambda}{L}}\approx \sqrt{\frac{\lambda}{2 L}},
\end{align}
where $\lambda$ is the emitted wavelength and $L=N\Lambda_u$ is the undulator length, with $\Lambda_u$ as the magnetic period of the undulator and $N$, the number of magnetic periods. The electron beam usually presents different horizontal and vertical beam emittances, but the undulator natural source size or divergence presents radial symmetry at the central cone (excluding polarization). The photon source size and divergence are approximately given by the convolution of the electron beam sizes and the undulator natural sizes. If we consider Gaussian distributions, the photon source is
\begin{align}
\label{eq:photon big sigmas}
\Sigma_{h,v}=\sqrt{\sigma_{e(h,v)}^2 + \sigma_u^2} && \Sigma_{h,v}'=\sqrt{\sigma_{e(h,v)}'^2 + \sigma_u'^2}.
\end{align}

The beam sizes at different positions in the beamline are estimated in different ways depending upon whether the beam is focused or not. If the beam is focused, its dimensions, to a first approximation, are the photon source size multiplied by the corresponding magnification factor. For predicting the photon beam sizes in a position out of the focus the beam divergence is used. In an ideal {\'{e}}tendue conserving optical system, where the Smith-Helmholz invariant applies, the resulting beam divergence transforms inversely as the beam size: it can be calculated as the product between the beam divergence at the source with the demagnification factor $M^{-1}$.

Using the values for U18.3, $n$=1, and  $E=17225$ eV (Table \ref{tab:Undulators}) and the optical configuration described in Table \ref{tab:Positions}, one obtains a focal size 139$\times$5 nm$^2$ (H$\times$V) for ESRF-1 and 10$\times$5 nm$^2$ for EBS (Table~\ref{tab:HandCalculations}). The horizontal focal spot for the ESRF-1 lattice is larger than the specified target value for the beamline, therefore the VSS could be closed to 50$\mu$m thus reducing the beamsize at the VSS plane by a factor ca. 8.
%\inred{8 $\approx$ 403.8/50=8.076}
Therefore, for ESRF-1, the horizontal size at the sample will be affected by the same factor obtaining roughly 17$\times$5 nm$^2$

In general, for simple elliptical cylinder figured grazing incident optics imaging extended sources with large demagnifications ($M^{-1} \gg 100$), the effect of optical aberrations is very important even using a perfect optical surface. As a consequence, the results calculated using only the demagnification factor are usually optimistic. Moreover imperfections in the optical surface (figure errors, slope errors and microroughness) are often the limiting factors of real reflective optics. Their effects will studied in Sections~\ref{level1} and \ref{level2}.
%Effects of slope errors and diffraction limits can be simply estimated by hand calculations as well...
%The smearing of the focal size because of diffractive effects by the mirror clipping will also be discussed.

\begin{table}\label{tab:HandCalculations}
    \centering
    \caption{Photon beam size and divergence for selected ID16A beamline positions. Values below were obtained considering undulator U18.3 tuned to its first harmonic at the photon energy of $E=17225$ eV for both the high-$\beta$ (ESRF) and the EBS straight sections. Values are FWHM. Values in parentheses correspond to closing the VSS slit to 50$\mu$m (21.3 if approximating by the Gaussian $\sigma$).}
    \resizebox{\textwidth}{!}{%
    \begin{tabular}{r|ccc|ccc}
                            &               & \textbf{ESRF (high-$\beta$)} &   &         & \textbf{EBS}  &    \\ 
    Element:                & Undulator     & VSS          & Sample            & Undulator & VVS   & Sample       \\
    Position [m]:           & 0.00          & 40.00        & 185.00            & 0.00    & 40.00   & 185.00       \\
    $FWHM_h$ [$\mu$m]:      & 977.19        & 404.00 (50.0)& 0.139 (0.017)     & 71.26   & 29.46   & 0.0102       \\
    $FWHM_v$ [$\mu$m]:      & 9.60          & 482          & 0.0052            & 10.02   & 486.5   & 0.0054       \\
    \newline
    $FWHM'_h$ [$\mu$rad]:   & 26.96         & 65.22        & 189062   & 15.60   & 37.74   & 109404       \\
    $FWHM'_v$ [$\mu$rad]:   & 12.04         & 12.04                 & 22264    & 12.16   & 12.16   & 22489        \\                                                   
    \end{tabular}%
    }
\end{table}

>>>>

\begin{synopsis}
A hierarchical scheme of the calculations of an X-ray beamline for coherent applications is presented. Starting from simple calculations to partial coherence calculations, we show how ray-tracing and wave optics software are applied simulating the beamline performances.   
\end{synopsis}

\begin{abstract}
We consider different approaches to simulating a modern X-ray beamline. Several methodologies with increasing complexity are applied to discuss the relevant parameters that quantify the beamline performance. Parameters such as flux, dimensions and intensity distribution of the focused beam and coherence properties are obtained from simple analytical calculations to sophisticated computer simulations using ray-tracing and wave optics techniques. A latest-generation X-ray nanofocusing beamline for coherent applications (ID16A at the ESRF) has been chosen to study in detail the issues related to highly demagnifying synchrotron sources and exploiting the beam coherence. The performance of the beamline is studied for two storage rings: the old ESRF-1 (emittance 4000~pm) and the new ESRF-EBS (emittance 150~pm). In addition to traditional results in terms of flux and beam sizes, an innovative study on the partial coherence properties based on the propagation of coherent modes is presented. The different algorithms and methodologies are implemented in the software suite OASYS \cite{codeOASYS}. Those are discussed with emphasis placed upon the their benefits and limitations.
\end{abstract}

@@

Many storage-ring-based X-ray synchrotron facilities are building or planning upgrades to increase brilliance and coherent flux by one to three orders of magnitude.  The first upgrade of a large facility will be the EBS (Extremely Brilliant Source) \cite{orangebook} at the European Synchrotron Radiation Facility (ESRF). The new storage ring of 150~pm emittance will be commissioned in 2020 and will significantly boost the coherence of the X-ray beams. Applications exploiting beam coherence such as X-ray photon correlation spectroscopy, coherent diffraction imaging, propagation-based phase-contrast imaging, and ptychography will strongly benefit from this update.

Accurate calculation and quantitative evaluation of the parameters related to X-ray optics are of great importance for designing, building and exploiting new beamlines. Every modern beamline follows a procedure of conceptual design plus technical design where detailed simulations are essential. These calculations allowing testing of the design parameters in a virtual computer environment where advantages and limitations can be studied quantitatively. Selection of the optimal optical layout requires the detailed simulation of the optics imperfections, that in many cases are the limiting factor of the beamline optical configuration. Deformations of the optical elements due to heat load need to be controlled, therefore optics simulations have to include results from engineering modeling of the thermal deformations usually done using finite element analysis. Traditionally the optics calculations for synchrotron beamlines are done using ray-tracing techniques. They model the X-ray beam as a collection of incoherent rays. This technique is still essential for obtaining information on aberrations, flux propagation, monochromator bandwidth, etc. However, the high coherence of the new sources can introduce the necessity to complement the ray-tracing methods with models including effects of diffraction and scattering from a coherent beam. A fully coherent hard X-ray beam is still a dream for the new generation storage rings. Despite a considerable increase in the coherent fraction of the new beams, the coherent fraction at typically hard X-ray photon energies is still a few percents for the new sources. Ray-tracing analysis does not become obsolete but has to be complemented with an analysis based on wave optics. Because of the partial coherence, the wave optics analysis becomes nontrivial and can be done with different levels of approximation, as discussed in this work. In addition, wave optics simulations are usually computationally expensive.

@@

X-ray beamlines are very particular optical systems. Some implications of exploiting the short wavelengths such as the use of grazing incidence reflectors and the use of perfect crystals as typical hard X-ray monochromators render common commercial software packages ill-adapted for the study of synchrotron beamlines. For this reason, many optics tools have been developed by the synchrotron community. Some of them have a long history like SHADOW - from 1984 \cite{firstShadow}; or SRW - from 1998 \cite{codeSRW}; to mention only codes used in this work. These historical tools together with other new applications have been collected together in a user-friendly software package: OASYS \cite{codeOASYS}. Many of the calculations presented here have been performed directly in OASYS or with scripts created with the help of OASYS.

 In this paper, some quantitative results on how the storage ring upgrade will affect a particular beamline are presented and it is structured as follows: we first introduce in Section~\ref{Beamline description} the main parameters of the beamline and explain the particular choice of the parameters used in the calculations. Section~\ref{Simulations} is the central part of the paper, and describes the different methods used to analyze the beamline with increasing complexity, setting a hierarchy of methodologies that can be exported to the simulations of any beamline. In Section \ref{level0} we calculate with simple methods approximate values of focal size, flux and coherent fraction. Accuracy can be improved using geometrical methods in Section \ref{level1} that allow a more complete simulation of the beamline optics within certain limits which are related, in this example, to high demagnification and high coherence of the studied beamline. The wave optics section~\ref{level2} includes different models of increasing complexity as the required accuracy of the results becomes more demanding. In the Discussion (Section \ref{discussion}) the results of the different simulation methods are commented. The hierarchical approach allows more accurate and better results but the price to pay is to deal with more complex and computationally expensive tools. Finally, Section \ref{summary} summarizes the results.

@@

This section presents the reader with the main operational parameters for the ID16A beamline at ESRF \cite{ID16A} in two scenarios. This beamline was built during the Upgrade Phase~I at ESRF (2009-2015). It uses one or several undulators as the source. During the period 2011-2018, it operated exploiting the ESRF-1 storage ring’s double-bend achromat (DBA) lattice with horizontal emittance about 4000~pm. In the second phase of the ESRF Upgrade, the original storage ring is being replaced by a new hybrid multi-bend achromat (HMBA) lattice. This new lattice will drastically improve the performances of the ESRF source. The resulting Extreme Brilliant Source (EBS) \cite{orangebook} will increase the brightness and coherent fraction by a factor of more than 100, by strongly reducing the horizontal emittance. This is possible using almost the same ring circumference and electron energy $E_e$. As a result, the X-ray sources will be much more coherent, especially in the horizontal direction.

ID16A is a beamline that provides a nanometric, highly coherent beam with broad energy-bandpass. It is dedicated to nanofocusing applications. The beamline can operate at two fixed photon energies: 17050~eV and 33600~eV. The straight section in the magnet lattice houses four undulators, two 18~mm and another two of 22.4~mm period. The undulator parameters and $K$-values at the resonances for the energies in use are shown in Table \ref{tab:Undulators}.

\begin{table}\label{tab:Undulators}
\centering
\caption{Parameters of the undulators available at the ID16A beamline: period ($\lambda_u$), length ($L$), number of identical undulators ($N$) and deflection parameter ($K$) values for the three main photon energies in use. The valuse are shown for the two ESRF magnetic lattices: the old ESRF-1 (labelled ESRF, with $E_e = 6.04$~GeV) and the new ESRF-EBS (labelled EBS, with $E_e = 6.00$~GeV). $n$ is the emission harmonic in use for each particular photon energy.}

\resizebox{\textwidth}{!}{%
\begin{tabular}{lcccccc}
name  & $\lambda_u$ [mm] & $L$ [m] & N & $K$ (11200~eV) & $K$ (17050~eV) & $K$ (33600~eV) \\ 
\input{tableUndulator.txt}
\end{tabular}%
}
\end{table}

ID16A is a long beamline: the distance from the center of the straight section to the sample position is 185~m. Strong focusing is performed using a Kirkpatrick-Baez (KB) mirror system placed a few cm upstream of the sample position (focal plane), which ensures a high demagnification. In order to match the acceptance of the KB in the vertical and horizontal planes, a focusing (cylindrically bent) multilayer monochromator (ML) has
been chosen. It i) monochromatizes the beam with a large energy bandwidth (typically $\Delta E/E \approx 10^{-2}$) to allow high flux, 
and ii) it focus tangentially the beam in the horizontal plane to create a secondary source at 40~m downstream of the undulator, where a horizontal slit (VSS) is placed. A beamline schematic is shown in Figure~\ref{fig:ID16A}. The positions of the different elements are shown in Table \ref{tab:Positions}. This table also shows the extremely high geometrical demagnification ($\sim 10^3$) needed for focusing the beam to the nanometric range.

@@

% Simulations were done using the two different storage ring lattices mentioned before: the (already dismantled) ESRF-1 where the undulators are in a high-$\beta$ straight section (larger horizontal emittance) and the new EBS (lower horizontal emittance) straight sections. The parameters for the current and the new lattice are shown in Tables \ref{tab:TableSources} and \ref{tab:eBeam}.

\begin{table}\label{tab:TableSources}
    \caption{Main parameters of the present DBA lattice (ESRF-1) and future HMBA lattice (EBS) storage rings \cite{orangebook}.}
        \begin{tabular}{rcc}
                                              & \textbf{ESRF-1} & \textbf{EBS}      \\ 
        Lattice type:                         & DBA           & HMBA              \\
        Circunference {[}m{]}:                & 844.930       & 843.979           \\
        Beam energy {[}GeV{]}:                & 6.04          & 6.00              \\
        Beam current {[}mA{]}:                & 200           & 200               \\
        Natural emittance {[}pm$\cdot$rad{]}: & 4000          & 147               \\
        Energy spread {[}$\%${]}:             & 0.0011        & 0.00093           \\       
    \end{tabular}
\end{table}

\begin{table}\label{tab:eBeam}
    \centering
    \caption{ESRF-1 and EBS main electron beam parameters. Electron beam parameters for the high-$\beta$ (labelled ESRF) and the EBS straight sections. Values taken at the symmetry point of the straight section, where the insertion devices are placed. Values for ESRF-1 are taken from \cite{ESRF2014}. For EBS, two (very close) design values are used, from Ref.~\cite{ESRF2014} labelled EBS-S28A, and those coming from further lattice refinement, labelled EBS-S28D. \todo{Remove ESRF-AT and EBS-S28A} }
        \resizebox{\textwidth}{!}{%
        \begin{tabular}{ccccc}
         Lattice & $\sigma_e$ horizontal {[}$\mu$m{]} &  $\sigma_e'$ horizontal {[}$\mu$rad{]} & $\sigma_e$ vertical {[}$\mu$m{]} & $\sigma_e'$ vertical {[}$\mu$rad{]} \\
         \input{tableElectron.txt}

@@

For all simulations presented hereafter, we selected the configuration using a single undulator U18.3 placed in the center of the straight section and tuned to its first harmonic at a photon energy of $E=17225$~eV: a value close to the usual operation value shown in Table \ref{tab:Undulators}. Calculations were done using the two different storage ring lattices mentioned before: the (already dismantled) ESRF-1 where the undulators are in a high-$\beta$ straight section (larger horizontal emittance) and the new EBS (lower horizontal emittance) straight sections. The parameters for the current and the new lattice are shown in Tables \ref{tab:TableSources} and \ref{tab:eBeam}.
This section is divided into three subsections, that present three main levels of simulations: simple calculations that can be made by hand, ray-tracing simulations and wave optics simulations: sections  \ref{level0}, \ref{level1} section \ref{level2} respectively.

@@

We first estimate the beam size at different positions in the beamline, in particular at the source plane and at the image plane. A first na{\"{i}}ve estimation of the focal size can be made by considering the source size times the magnification factor $M$. In the paraxial approximation, the magnification factor can be expressed as $M=q/p$, where $p$ is the distance between object and imaging element and $q$ is the distance between the imaging element and image. The demagnification is the inverse of the magnification ratio: $M^{-1}=p/q$. Systems where $M>1$ are said to be magnifying, whereas systems with $M<1$ (or $M^{-1}>1$) are said to be demagnifying. The latter will be our case: it is, consequently, more convenient to work with demagnification ($M^{-1}>1$) rather than with magnification.

The source size and divergence are usually approximated supposing a Gaussian distribution of the electron positions and divergences (Table \ref{tab:eBeam}) and single-electron photon source size and divergence. In some cases, such as for undulator emission, this Gaussian approximation may not be accurate and have some implications that will be discussed later. Let $\sigma_{e(h,v)}$ and $\sigma_{e(h,v)}'$ be the standard deviations of electron beam size and divergence at the center of symmetry of the straight section in a plane perpendicular to the main propagation direction. The subscripts $h$ and $v$ stand for horizontal and vertical, respectively.

The undulator natural photon source size and divergence are given by \cite{elleaume}:
\begin{align}
    \label{eq:photon small sigmas}
    \sigma_u=\frac{2.74}{4\pi}\sqrt{\lambda L}\approx \frac{1}{2  \pi}\sqrt{2 \lambda L}  && \sigma_u' = 0.69\sqrt{\frac{\lambda}{L}}\approx \sqrt{\frac{\lambda}{2 L}},
\end{align}
where $\lambda$ is the emitted wavelength and $L=N\Lambda_u$ is the undulator length, with $\Lambda_u$ as the magnetic period of the undulator and $N$, the number of magnetic periods. The electron beam usually presents different horizontal and vertical beam emittances, but the undulator natural source size or divergence presents radial symmetry at the central cone (excluding polarization). The photon source size and divergence are approximately given by the convolution of the electron beam sizes and the undulator natural sizes. If we consider Gaussian distributions, the photon source is:
\begin{align}
\label{eq:photon big sigmas}
\Sigma_{h,v}=\sqrt{\sigma_{e(h,v)}^2 + \sigma_u^2} && \Sigma_{h,v}'=\sqrt{\sigma_{e(h,v)}'^2 + \sigma_u'^2}.
\end{align}

The beam sizes at different positions in the beamline are estimated in different ways depending upon whether the beam is focused or not. If the beam is focused, its dimensions, to a first approximation, are the photon source size multiplied by the corresponding magnification factor. For predicting the photon beam sizes in a position out of the focus the beam divergence is used. In an ideal {\'{e}}tendue-conserving optical system, where the Smith-Helmholz invariant applies, the resulting beam divergence transforms inversely as the beam size: it can be calculated as the product between the beam divergence at the source with the demagnification factor $M^{-1}$.

Using the values for U18.3, $n=1$, and  $E=17225$~eV (Table \ref{tab:Undulators}) and the optical configuration described in Table \ref{tab:Positions}, one obtains a focal size 139$\times$5~nm$^2$ (H$\times$V) for ESRF-1 and $10\times5$~nm$^2$ for EBS (Table~\ref{tab:HandCalculations}). The horizontal focal spot for the ESRF-1 lattice is larger than the specified target value for the beamline, therefore the VSS could be closed to 50~$\mu$m thus reducing the beam size at the VSS plane by a factor ca. 8.
%\inred{8 $\approx$ 403.8/50=8.076}
Therefore, for ESRF-1, the horizontal size at the sample will be affected by the same factor obtaining roughly 17$\times$5~nm$^2$.

In general, for simple elliptical cylinder figured grazing incident optics imaging extended sources with large demagnifications ($M^{-1} \gg 100$), the effect of optical aberrations is very important even using a perfect optical surface. As a consequence, the results calculated using only the demagnification factor are usually optimistic. Moreover, imperfections in the optical surface (figure errors, slope errors and microroughness) are often the limiting factors of real reflective optics. Their effects will be studied in Sections~\ref{level1} and \ref{level2}.

%Effects of slope errors and diffraction limits can be simply estimated by hand calculations as well...
%The smearing of the focal size because of diffractive effects by the mirror clipping will also be discussed.

\begin{table}\label{tab:HandCalculations}
    \centering
    \caption{Photon beam size and divergence for selected ID16A beamline positions. Values below were obtained considering undulator U18.3 tuned to its first harmonic at the photon energy of $E=17225$~eV for both the high-$\beta$ (ESRF-1) and the EBS straight sections. Values are FWHM. Values in parentheses correspond to closing the VSS slit to 50~$\mu$m (21.3 if approximating by the Gaussian $\sigma$).}
    \resizebox{\textwidth}{!}{%
    \begin{tabular}{r|ccc|ccc}
                            &               & \textbf{ESRF (high-$\beta$)} &   &         & \textbf{EBS}  &    \\ 
    Element:                & Undulator     & VSS          & Sample            & Undulator & VVS   & Sample       \\
    Position [m]:           & 0.00          & 40.00        & 185.00            & 0.00    & 40.00   & 185.00       \\
    FWHM$_h$ [$\mu$m]:      & 977.19        & 404.00 (50.0)& 0.139 (0.017)     & 71.26   & 29.46   & 0.0102       \\
    FWHM$_v$ [$\mu$m]:      & 9.60          & 482          & 0.0052            & 10.02   & 486.5   & 0.0054       \\
    \newline
    FWHM'$_h$ [$\mu$rad]:   & 26.96         & 65.22        & 189062   & 15.60   & 37.74   & 109404       \\
    FWHM'$_v$ [$\mu$rad]:   & 12.04         & 12.04                 & 22264    & 12.16   & 12.16   & 22489        \\                                                   
    \end{tabular}%
    }
\end{table}

Next, we want to estimate the number of photons in the sample. For that, we need the spectrum at the source. It is not straightforward to get this information analytically but one can calculate the flux and power using different codes. Perhaps the most advanced and well-maintained codes used in the synchrotron community are SPECTRA \cite{codeSPECTRA} and SRW \cite{codeSRW}. The XOPPY package in the OASYS suite implements three codes: US \cite{codeUS}, URGENT \cite{codeURGENT} and SRW. Figure~\ref{fig:FluxU18} shows the spectra calculated using SRW over an aperture of $1 \times 1$~mm$^2$ placed at $d=27.1$~m. This aperture opening is set to fully accept the central cone at the resonance peak of the harmonic in use ($\max(\Sigma') \times d = 0.73$~mm), with $\Sigma'$ values from in Table~\ref{tab:HandCalculations}. The peak intensity is about  $2.9\cdot10^{14}$ photons/s/0.1{\%}bw for EBS and $1.9\cdot10^{14}$ for ESRF-1 (see \ref{fig:FluxU18}). One can appreciate how the emittance plays an important role in both peak width and especially in peak value. The number of photons per second after the multilayer monochromator is $2.9\cdot10^{15}$ for EBS and $1.9\cdot10^{15}$ for ESRF-1, considering that the energy bandwidth transmitted by a multilayer monochromator is $\Delta E/E \approx 10^{-2}$ and the bandwidth in Fig.~\ref{fig:FluxU18} is 10$^{-3}$, and supposing, at this point, ideal reflectivity (100~\%) for the multilayers.
